How topos theory can help algebra and geometry

### 2018-07-10 slides-como2018.tex
typo

  More details on this research program can be found in
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
  notes}, partly reported on at the
  \fixedhref{https://www.youtube.com/watch?v=7S8--bIKaWQ}{2015 IHÉS
  conference}. Even though many important dictionary entries
  are still missing (for instance pertaining derived categories and
  intersection theory), I believe that it is already at its current stage
  useful to working algebraic geometers.

>>>>

  More details on this research program can be found in
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
  notes}, partly reported on at the
  \fixedhref{https://www.youtube.com/watch?v=7S8--bIKaWQ}{2015 IHÉS
  conference}. Even though many important dictionary entries
  are still missing (for instance pertaining to derived categories and
  intersection theory), I believe that it is already at its current stage
  useful to working algebraic geometers.

### 2018-07-10 slides-leipzig2018.tex
typo

\newcommand{\bignumber}[1]{
  \renewcommand{\insertenumlabel}{#1}\scalebox{1.5}{\usebeamertemplate{enumerate item}}
}

@@

\newcommand{\bad}[1]{\textcolor{red!90}{\textnormal{#1}}}

@@

  Crucially, two theories~$T$ and~$T'$ can have equivalent classifying
  toposes even when they are not syntactially related in any way. This
  observation is the starting point of Olivia Caramello's
  \fixedhref{https://www.oliviacaramello.com/}{\emph{bridge
  technique}}, a grand research program with applications in many different
  fields.

@@

  For any topological space~$X$, there is the topos~$\Sh(X)$ of \emph{sheaves
  over~$X$}. They are useful in analysis for internalizing
  parameter-dependence. Apart from pathological cases like~$X$
  being discrete, the intermediate value theorem in the form
  \begin{indentblock}Let~$f : \RR \to \RR$ be a continuous function.
  Assume~$f(-1) < 0 < f(1)$. Then there is a number~$x \in \RR$ such
  that~$f(x) = 0$.\end{indentblock}
  fails in these toposes -- for very meaningful reasons discussed below.
  The following, classically equivalent, version does hold:
  \begin{indentblock}Let~$f : \RR \to \RR$ be a continuous function.
  Assume~$f(-1) < 0 < f(1)$. Then, for every~$\varepsilon > 0$, there is
  a number~$x \in \RR$ such that~$|f(x)| < \varepsilon$.\end{indentblock}

@@

  \jnote{1-2}{We can picture the Kripke--Joyal semantics as a kind of translation
  engine. When we're talking with Anna, a mathematician who lives in the
  ef{}fective topos, at first we might be weirded out when she states ``it's a basic
  fact of life that any function~$\NN \to \NN$ is computable''. But if we
  remember to switch on the Kripke--Joyal translation, we instead hear~``it's a
  basic fact of life that there is a Turing machine which, given a Turing
  machine computing a function~$f : \NN \to \NN$, outputs a Turing machine
  computing~$f$'' which we can easily agree with.

@@

  A recommendation for more on constructive mathematics is the informative and
  entertaining talk \emph{Five Stages of Accepting Constructive Mathematics}
  by Andrej Bauer
  (\fixedhref{https://www.youtube.com/watch?v=21qPOReu4FI}{video},
  \fixedhref{http://www.ams.org/journals/bull/2017-54-03/S0273-0979-2016-01556-4/}{notes}).}

@@

  \jnote{1}{The applications of the internal language of toposes in several
  branches of mathematics are driven by the following fundamental empirical fact:
  \begin{center}The external translations of \\
  some easy-to-prove internal statements of suitable toposes \\
  are \\
  nontrivial statements about the subject matter at hand.\end{center}
  For instance, Grothendieck's generic freeness lemma, an important theorem in
  commutative algebra and algebraic geometry, is the external translation of
  the basic observation ``finitely generated modules over finitely generated
  algebras over fields are \emph{not not} free over the field and \emph{not
  not} finitely presented over the algebra'' of constructive linear algebra.

@@

  The next slide shows two examples.}

  \jnote{2}{\begin{itemize}\justifying
    \item The statement ``if the two outer sheaves in a short exact sequence of
    sheaves of modules are of finite type, then so is the middle one'' is the
    external translation of the basic fact ``if the two outer modules
    in a short exact sequence of modules are finitely generated, then so is the
    middle one'' of constructive linear algebra.

    The syntactical complexity of the internal statement is quite low (just one
    implication sign, on toplevel, no long chains of quantifiers of mixed
    types). Therefore the resulting external proof obtained by unwinding will
    still be quite short and straightforward. In fact, this external proof will
    coincide with what anyone well-versed in scheme theory will produce,
    judiciously juggling with open subsets.

    Using the internal language of toposes here is therefore mostly for mental
    hygiene, making rigorous the intuitive idea that the two statements are
    related to each other.

    \item A version of Hilbert's basis theorem, ``polynomial rings over
    anonymously Noetherian rings are again anonymously Noetherian'', can be put
    to use in the proof of Grothendieck's generic freeness lemma alluded to
    above. This statement has nontrivial syntactical complexity (the
    Noetherian condition involves a double negation, therefore nested
    implications). For this reason, already the translation of the
    \emph{statement} is quite convoluted (a part of it is
    \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-como2018.pdf}{displayed
    in this set of slides} for your viewing pleasure). The translation of its
    \emph{proof} contains a deformed copy of the standard proof of Hilbert's
    basis theorem (in a way which doesn't allow to simply \emph{apply}
    Hilbert's basis theorem, one has to actually redo its proof).

    In situations like this, the internal language is of real value.
  \end{itemize}}

  \jnote{3}{We're used to the fact that the usual laws of mathematical
  reasoning apply to mathematical objects. The internal language of toposes
  encapsulates and makes useful the following observation: The usual laws of
  reasoning apply to our favorite mathematical objects \emph{also in
  nontrivial, different ways}. I find this quite astonishing.

  The next slide shows a basic application of this train of thought in
  analysis. The presented example is of a rather simple nature and serves only
  to explain the translation process in an explicit and rigorous fashion,
  contrasting the previous slides which were somewhat short on details.
  Details on the application to homotopy theory can be found in the
  \fixedhref{https://homotopytypetheory.org/book/}{HoTT book}.}
\end{frame}

@@

  Synthetic differential geometry was born to provide an account of
  differential geometry which is both rigorous and closer to intuition than the
  standard account. Its starting point is the axiom of microaffinity, which is
  very useful but wildly false in classical mathematics. The fundamental
  theorem about synthetic differential geometry, connecting it with the
  ordinary world of smooth manifolds, is that there exist \emph{well-adapted
  models} for synthetic differential geometry -- toposes~$\E$ such that:
  \begin{enumerate}\justifying
    \item There is a functorial way of associating to any smooth manifold~$M$
    an object~$y(M)$ of~$\E$ (something which~$\E$ believes to be a set) and
    to any smooth map~$f : M \to N$ a morphism~$y(f) : y(M) \to y(N)$ of~$\E$
    (something which~$\E$ believes to be map).
    % (item is referenced below by number)
    \item Any morphism in~$\E$ of type~$y(M) \to y(N)$ is
    of the form~$y(f)$ for a smooth map~$f$, and if~$\E \models y(f) = y(g)$,
    then actually~$f = g$.
    \item In~$\E$, the axiom of microaffinity and related axioms hold for the
    ring~$y(\RR^1)$.
  \end{enumerate}}

@@

  \jnote{3}{Recall the following standard reflection principle taught in
  undergraduate calculus: Even though continuity is officially a property which
  depends only on the input/output behaviour of a function, reflecting on the
  \emph{term} used to define a given function often allows us to quickly
  conclude that it's continuous. For instance, any function defined by a
  polynomial expression is continuous.

@@

  Synthetic differential geometry is very well-developed and at your disposal.
  References include:
  \begin{itemize}
    \item \emph{\fixedhref{https://rawgit.com/iblech/mathezirkel-kurs/master/thema05-sdg/blatt05.pdf}{Synthetische
    Differentialgeometrie}}, notes for high school students (in German)
    \item \emph{\fixedhref{http://math.andrej.com/2008/08/13/intuitionistic-mathematics-for-physics/}{Intuitionistic
    mathematics for physics}}, a blog post by Andrej Bauer
    \item \emph{A Primer of Infinitesimal Analysis}, a short introduction by John Bell
    \item \emph{\fixedhref{http://home.imf.au.dk/kock/sdg99.pdf}{Synthetic
    Differential Geometry}}, the definitive book by Anders Kock
  \end{itemize}

  [Technical comment: The object~$y(\RR^1)$ will usually not be the object of
  Dedekind real numbers in~$\E$.]}
\end{frame}

@@

  \jnote{3}{As mentioned before, any proof involving the internal language can be
  \emph{unwound} to yield a direct proof not referencing toposes. Depending on
  the logical complexity of the statements appearing in the internal proof,
  this process can substantially lengthen the proof. In the case of the new
  proof of Grothendieck's generic freeness lemma, we were lucky to
  \fixedhref{https://arxiv.org/abs/1807.01231}{obtain a one-page proof} using
  this process. (That the resulting external proof was so short is because it
  was possible to eliminate the use of the Noetherian property from the
  internal proof; without that elimination, the unwound proof was much longer.)}
\end{frame}

@@

  \jnote{2}{This tension is resolved by observing that the category of
  functors~$\Ring \to \Set$ is a topos (the \emph{big Zariski topos} of~$\Spec(\ZZ)$)
  and that we can therefore employ its internal language. This language takes
  care of juggling stages behind the scenes. For instance, projective~$n$-space
  can be described by the naive expression
  \[ \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\, x_0 \neq 0 \vee \cdots \vee x_n
  \neq 0 \}/(\affl)^\times. \]
  This example illustrates the goal: to develop a synthetic account of
  algebraic geometry, in which schemes are plain sets and morphisms between
  schemes are maps between those sets. It turns out that there are many
  similarities with the well-developed synthetic account of differential
  geometry, but also important differences, and it also turns out that
  synthetic algebraic geometry has close connections to a certain age-old
  unresolved problem in topos theory, the \emph{mystery of nongeometric
  sequents}.

>>>>

@@

@@

  Crucially, two theories~$T$ and~$T'$ can have equivalent classifying
  toposes even when they are not syntactically related in any way. This
  observation is the starting point of Olivia Caramello's
  \fixedhref{https://www.oliviacaramello.com/}{\emph{bridge
  technique}}, a grand research program with applications in many different
  fields.

@@

  For any topological space~$X$, there is the topos~$\Sh(X)$ of \emph{sheaves
  over~$X$}. They are useful in analysis for internalizing
  parameter-dependence. Apart from pathological cases like~$X$
  being discrete, the intermediate value theorem in the form
  \begin{indentblock}\emph{Let~$f : \RR \to \RR$ be a continuous function.
  Assume~$f(-1) < 0 < f(1)$. Then there is a number~$x \in \RR$ such
  that~$f(x) = 0$.}\end{indentblock}
  fails in these toposes -- for very meaningful reasons discussed below.
  The following, classically equivalent, version does hold:
  \begin{indentblock}\emph{Let~$f : \RR \to \RR$ be a continuous function.
  Assume~$f(-1) < 0 < f(1)$. Then, for every~$\varepsilon > 0$, there is
  a number~$x \in \RR$ such that~$|f(x)| < \varepsilon$.}\end{indentblock}

@@

  \jnote{1-2}{We can picture the Kripke--Joyal semantics as a kind of translation
  engine. When we're talking over the phone with Anna, a mathematician who lives in the
  ef{}fective topos, at first we might feel uncomfortable when she states ``it's a basic
  fact of life that any function~$\NN \to \NN$ is computable''. But if we
  remember to switch on the Kripke--Joyal translation, we instead hear~``it's a
  basic fact of life that there is a Turing machine which, given a Turing
  machine computing a function~$f : \NN \to \NN$, outputs a Turing machine
  computing~$f$'' which we can easily agree with.

@@

  A recommendation for more on constructive mathematics is the informative and
  entertaining talk \emph{Five Stages of Accepting Constructive Mathematics}
  by Andrej Bauer
  (\fixedhref{https://www.youtube.com/watch?v=21qPOReu4FI}{video},
  \fixedhref{https://www.ams.org/journals/bull/2017-54-03/S0273-0979-2016-01556-4/}{notes}).}

@@

  \jnote{1}{The applications of the internal language of toposes in several
  branches of mathematics are driven by the following fundamental empirical fact:
  \vspace*{-1em}
  \begin{center}The external translations of \\
  some easy-to-prove internal statements of suitable toposes \\
  are \\
  nontrivial statements about the subject matter at hand.\end{center}
  \vspace*{-1em}
  For instance, Grothendieck's generic freeness lemma, an important theorem in
  commutative algebra and algebraic geometry, is the external translation of
  the basic observation ``finitely generated modules over finitely generated
  algebras over fields are \emph{not not} free over the field and \emph{not
  not} finitely presented over the algebra'' of constructive linear algebra.

@@

  The next two slides show examples regarding this increase in complexity.}

  \jnote{2}{\begin{enumerate}\justifying
    \item The statement ``if the two outer sheaves in a short exact sequence of
    sheaves of modules are of finite type, then so is the middle one'' is the
    external translation of the basic fact ``if the two outer modules
    in a short exact sequence of modules are finitely generated, then so is the
    middle one'' of constructive linear algebra.

    The syntactical complexity of the internal statement is quite low (just one
    implication sign, at toplevel, no long chains of quantifiers of mixed
    types). Therefore the resulting external proof obtained by unwinding will
    still be quite short and straightforward. In fact, this external proof will
    coincide with what anyone well-versed in scheme theory will produce,
    judiciously juggling open subsets.

    Using the internal language of toposes in situations like this is therefore
    mostly for mental hygiene, making rigorous the intuitive idea that the two
    statements are related to each other.
  \end{enumerate}}

  \jnote{3}{\begin{enumerate}\justifying\addtocounter{enumi}{1}
    \item A version of Hilbert's basis theorem, ``polynomial rings over
    anonymously Noetherian rings are again anonymously Noetherian'', can be put
    to use in the proof of Grothendieck's generic freeness lemma alluded to
    above.

    This statement has nontrivial syntactical complexity (the
    Noetherian condition involves a double negation, therefore nested
    implications). For this reason, already the translation of the
    \emph{statement} is quite convoluted (a part of it is
    \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-como2018.pdf}{displayed
    in this set of slides} for your viewing pleasure). The translation of its
    \emph{proof} contains a deformed copy of the standard proof of Hilbert's
    basis theorem (in a way which doesn't allow to simply \emph{apply}
    Hilbert's basis theorem, one has to actually redo its proof).

    In situations like this, the internal language is of real value.
  \end{enumerate}}

  \jnote{4}{We're used to the fact that the usual laws of mathematical
  reasoning apply to mathematical objects. The internal language of toposes
  encapsulates and makes useful the following observation: The usual laws of
  reasoning apply to our favorite mathematical objects \emph{also in
  nontrivial ways different from the accustomed one}. I find this quite astonishing.

  The next slide shows a basic application of this train of thought in
  analysis. The presented example is of a rather simple nature and serves only
  to explain the translation process in an explicit and rigorous fashion,
  contrasting the previous slides which were somewhat short on details.
  Details on the application to homotopy theory can be found in the
  \fixedhref{https://homotopytypetheory.org/book/}{HoTT book}.

  A tour of applications in algebraic geometry can be found in
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
  notes}.}
\end{frame}

@@

  Synthetic differential geometry was born to provide an account of
  differential geometry which is both rigorous and closer to intuition than the
  standard account, allowing for instance to read the classic works of Sophus
  Lie in a literal manner.

  Its starting point is the axiom of microaffinity, which is
  quite useful but wildly false in classical mathematics. The fundamental
  theorem about synthetic differential geometry, connecting it with the
  ordinary world of smooth manifolds, states that there exist \emph{well-adapted
  models} for synthetic differential geometry -- toposes~$\E$ such that:\par
  \scriptsize
  \begin{enumerate}\justifying
    \item There is a functorial way of associating to any smooth manifold~$M$
    an object~$y(M)$ of~$\E$ (something which~$\E$ believes to be a set) and
    to any smooth map~$f : M \to N$ a morphism~$y(f) : y(M) \to y(N)$ of~$\E$
    (something which~$\E$ believes to be map).
    % (item is referenced below by number)
    \item Any morphism of~$\E$ of type~$y(M) \to y(N)$ is
    of the form~$y(f)$ for a smooth map~$f$, and if~$\E \models y(f) = y(g)$,
    then actually~$f = g$.
    \item In~$\E$, the axiom of microaffinity and related axioms hold for the
    ring~$y(\RR^1)$. [Technical comment: This ring will usually not be the ring of
    Dedekind real numbers in~$\E$.]
  \end{enumerate}}

@@

  \jnote{3}{The following reflection principle is routinely used in
  undergraduate calculus: Even though continuity is officially a property which
  depends only on the input/output behaviour of a function, reflecting on the
  \emph{term} used to define a given function often allows us to quickly
  conclude that it's continuous. For instance, any function defined by a
  polynomial expression is continuous.

@@

  Synthetic differential geometry is very well-developed and at your disposal.
  References include:
  \begin{itemize}
    \item \emph{\fixedhref{https://rawgit.com/iblech/mathezirkel-kurs/master/thema05-sdg/blatt05.pdf}{Synthetische
    Differentialgeometrie}}, notes for high school students (in German)
    \item \emph{\fixedhref{http://math.andrej.com/2008/08/13/intuitionistic-mathematics-for-physics/}{Intuitionistic
    mathematics for physics}}, a blog post by Andrej Bauer
    \item \emph{A Primer of Infinitesimal Analysis}, a short introduction by John Bell
    \item \emph{\fixedhref{http://home.imf.au.dk/kock/sdg99.pdf}{Synthetic
    Differential Geometry}}, the definitive book by Anders Kock
  \end{itemize}}
\end{frame}

@@

  \jnote{3}{As mentioned before, any proof involving the internal language can be
  \emph{unwound} to yield a direct proof not referencing toposes. Depending on
  the logical complexity of the statements appearing in the internal proof,
  this process can substantially lengthen the proof.

  In the case of the new
  proof of Grothendieck's generic freeness lemma, we were lucky to
  \fixedhref{https://arxiv.org/abs/1807.01231}{obtain a one-page proof} using
  this process. (That the resulting external proof was so short is because it
  was possible to eliminate the use of the Noetherian property from the
  internal proof; without that elimination, the unwound proof was much longer.)}
\end{frame}

@@

  \jnote{2}{This tension is resolved by observing that the category of
  functors~$\Ring \to \Set$ is a topos (the \emph{big Zariski topos} of~$\Spec(\ZZ)$)
  and that we can therefore employ its internal language. This language takes
  care of juggling stages behind the scenes. For instance, projective~$n$-space
  can be described by the naive expression
  \[ \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\, x_0 \neq 0 \vee \cdots \vee x_n
  \neq 0 \}/(\affl)^\times. \]
  This example illustrates the goal: to develop a synthetic account of
  algebraic geometry, in which schemes are plain sets and morphisms between
  schemes are maps between those sets. It turns out that there are many
  similarities with the well-developed synthetic account of differential
  geometry, but also important differences, and it also turns out that
  synthetic algebraic geometry has close connections to a certain age-old
  unsolved problem in topos theory, the \emph{mystery of nongeometric
  sequents}.
